---
title: "Analysis: Differential Expression"
format: html
---

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)
```

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
# Differential expression
# Get full DE table
de_table_file <- dir(
  "analysis_de",
  pattern = "^.*\\.de\\.full\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)[1]

de_table <- read_csv(de_table_file) %>%
  mutate(
    pct.1 = pct.1 * 100,
    pct.2 = pct.2 * 100
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

sig_de_table <- de_table %>% filter(sig == "sig")

# Get cohort ID
cohort_id <- gsub("^analysis_de\\/(.*)\\.de\\.full\\.csv$", "\\1", de_table_file)

# Get all comparisons
all_comparisons <- unique(de_table$comparison)

# Get comparisons CSV
comparisons_csv <- "samplesheets/comparisons.csv"
comparisons_df <- read_csv(comparisons_csv)

# Order and reduce to necessary columns
de_table_col_order <- c(
  "Gene",
  "Gene.Symbol",
  "test_group",
  "ref_group",
  "pct.1",
  "pct.2",
  "avg_log2FC",
  "p_val",
  "p_val_adj_all",
  "neg_log10_pval_adj_all"
)
de_table_col_names_ordered <- c(
  "Gene ID",
  "Gene Symbol",
  "Test Group",
  "Reference Group",
  "% Cells Expressing in Test Group",
  "% Cells Expressing in Reference Group",
  "Average log2 Fold Change",
  "Raw p-value",
  "FDR-Adjusted p-value",
  "-log10 Adjusted p-value"
)

de_table <- de_table %>% select(all_of(de_table_col_order))
sig_de_table <- sig_de_table %>% select(all_of(de_table_col_order))

# Get volcano plots
volcano_plot_file <- paste0("analysis_de/results/", cohort_id, ".volcano.png")

# Get p- and log-fc- distribution plots
p_dist_plot_file <- paste0("analysis_de/results/", cohort_id, ".p_val_dist.png")
log_fc_dist_plot_file <- paste0("analysis_de/results/", cohort_id, ".log_fc_dist.png")
```

This page shows the differential expression results for the comparisons you requested on your dataset. The results are presented in four sections:

- [p-value distribution results](#p-value-distributions)
- [Log-fold change distribution results](#log-fold-change-distributions)
- [Volcano plots](#volcano-plots)
- [Differential expression results table](#differential-expression-table)

#### Dataset name: `{r} cohort_id`

#### All comparisons made:

You provided the following table of comparisons to perform via the `--comparisons` pipeline parameter:

```{r}
datatable(comparisons_df, colnames = c("Reference group", "Test group"))
```

*Note that the group names above must match the [pseudobulk groups shown in the previous section of the report](analysis_pseudo.qmd#pseudobulked-groups-availablle-for-comparison).*

### Differential expression results

#### p-value distributions

The following plots show the distribution of p-values for each comparison made. These can be useful to determine the quality of the results: a well-powered experiment with true positive results shoudl show an enrichment of low p-values (i.e. towards the left side of each plot). Uniformly distributed p-values indicate no true positive results or low power to detect true positives.

```{r}
#| results: asis
#| echo: false
cat(paste0('<img src="', p_dist_plot_file, '" alt="p-value distributions: ', cohort_id, '" width="100%">\n\n'))
```

#### Log-fold-change distributions

Log-fold-change distributions are also useful in determining **how much** your significant genes are differentially expressed by. In the plots below, the red histograms represent the distribution of non-significant ("ns") genes, while the blue histograms represent significant genes ("sig").

You can use these plots to determine whether you want to apply an additional fold-change threshold to your data. Such a threshold can help to remove noise from very lowly perturbed genes that may not represent a significant biological change. You can apply a fold-change threshold with the `--fc_threshold` parameter. **Note** that this parameter takes a **positive, non-log-transformed value**; i.e., to filter out all results with less than a 1.5x change in either direction (equivalent to ±0.58 in these log-scaled plots), you would provide `--fc_threshold 1.5`.

Keep in mind that in many biological contexts, small fold-changes can be very biologically significant, so only use this parameter if you are sure you want to discard low fold-change results.

```{r}
#| results: asis
#| echo: false
cat(paste0('<img src="', log_fc_dist_plot_file, '" alt="log-fold-change distributions: ', cohort_id, '" width="100%">\n\n'))
```

#### Volcano plots

Volcano plots combine both the p-value and fold-change information into one very useful plot. At a glance, you can identify the most significant results (higher up on the vertical axis, and highlighted in red) and the most differentially expressed genes (further away from the centre). If you have provided a fold-change threshold, vertical lines will be drawn at the relevant fold-change positions on the plot and any genes between them will be greyed out and considered non-significant. The plots below additionally highlight the top-most significant genes for your convenience.

```{r}
#| results: asis
#| echo: false
cat(paste0('<img src="', volcano_plot_file, '" alt="volcano plots: ', cohort_id, '" width="100%">\n\n'))
```

#### Differential expression table

Finally, below is a table summarising all of the differential expression results performed. The "Significant results only" tab shows just those genes that pass the p-value and fold-change thresholds provided. The "All results" tab shows all genes from all comparisons, regardless of significance.

:::panel-tabset

##### Significant results only

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
datatable(sig_de_table, colnames = de_table_col_names_ordered)
```

##### All results

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
datatable(de_table, colnames = de_table_col_names_ordered)
```

:::
