---
title: "Analysis: Over-Representation Analysis"
format: html
---

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)
```

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
# ORA
# Get full ORA table
ora_table_file <- dir(
  "analysis_ora",
  pattern = "^.*\\.ora\\.full\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)[1]
ora_table <- read_csv(ora_table_file)

# Get cohort ID
cohort_id <- gsub("^analysis_ora\\/(.*)\\.ora\\.full\\.csv$", "\\1", ora_table_file)

# Get reduced GSEA table
ora_reduced_table_file <- paste0("analysis_ora/", cohort_id, ".ora.full.reduced.csv")
ora_reduced_table <- read_csv(ora_reduced_table_file)

# Remove unnecessary columns
retain_cols <- c(
  "geneSet",
  "description",
  "test_group",
  "ref_group",
  "enrichmentRatio",
  "size",
  "overlap",
  "pValue",
  "FDR"
)
retain_cols_names <- c(
  "Gene Set ID",
  "Gene Set Description",
  "Test Group",
  "Reference Group",
  "Enrichment Ratio",
  "Gene Set Size",
  "# Input Genes in Set",
  "p-value",
  "FDR-Adjusted p-value"
)

ora_table <- ora_table %>%
  select(all_of(retain_cols)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))
ora_reduced_table <- ora_reduced_table %>%
  select(all_of(retain_cols)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

# Get all comparisons
unique_comparisons <- ora_table %>%
  select(test_group, ref_group) %>%
  unique()
all_comparisons <- unique_comparisons %>%
  unite(comparison, test_group, ref_group, sep = " vs. ") %>%
  .$comparison
all_comparisons_str <- unique_comparisons %>%
  unite(comparison, test_group, ref_group, sep = "_vs_") %>%
  .$comparison
names(all_comparisons_str) <- all_comparisons

# Get comparisons CSV
comparisons_csv <- "samplesheets/comparisons.csv"
comparisons_df <- read_csv(comparisons_csv)

# Get ORA plots
ora_plots <- sapply(all_comparisons_str, function(cmp) {
  paste0("analysis_ora/plots/", cohort_id, ".ora.", cmp, ".png")
})
```

Differential expression analyses can generate lots of results that can be quite difficult to interpret by themselves. Functional enrichment analyses are a set of methods designed to take the raw differential expression results and identify patterns of expression that point to altered **biological function**. There are several such approaches, but the two we apply in this pipeline are **Gene Set Enrichment Analysis** (GSEA) and **Over-Representation Analysis** (ORA).

This page presents the results from the ORA analysis conducted on your dataset. ORA works by taking the set of your **significantly differentially expressed** genes and comparing that against each set of genes mapped to **biological pathways** in a curated database. For each pathway, ORA looks at the proportion of those genes that appear in your list of significant genes. If that proportion is higher than what would be expected by chance, the gene set is considered **over-represented** or **enriched** in your dataset. In contrast, some pathways may have very low representation in your dataset, in which case they can be considered **under-represented** or **depleted**. The ORA method generates a list of **enrichment ratios** for each pathway.

#### Dataset name: `{r} cohort_id`

#### All comparisons made:

One GSEA analysis was conducted for each differential expression comparison that was made. As per the [differential expression page](analysis_de.qmd), these comparisons were:

```{r}
datatable(comparisons_df, colnames = c("Reference group", "Test group"))
```

### ORA expression results

Below, you will find plots for each ORA analysis. These plots show the log2-transformed enrichment ratio for each **significant** pathway from the analysis; more enriched pathways have bars pointing to the right and are highlighted in orange; depleted or under-represented pathways point to the left and are highlighted in blue.

:::panel-tabset

```{r}
#| results: asis
purrr::walk(
  sort(all_comparisons),
  function(n) {
    p <- ora_plots[[n]]
    cat(paste("####", n, "\n\n"))
    cat(paste("#####", "Over-Representation Analysis Results:", n))
    cat(paste0('<img src="', p, '" alt="ora results: ', n, '" width="100%">\n\n'))
  }
)
```

:::