---
title: "scRNAvigator Report"
format:
  html:
    embed-resources: true
---

Thank you for using scRNAvigator!

This report contains a summary of your analysis. Depending on what sections of the pipeline were ran, it will include one or more of the following:

- Quality control metrics and plots
- Doublet detection and filtering results
- Dataset integration results
- Cell annotation results
- Pseudobulking and differential expression results
- Functional enrichment analysis results

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)

# Get cohort name
cohort_name <- scan("cohort_name.txt", character())[1]

# Get parameters TSV file
params_tsv <- "parameters.tsv"
params_df <- read_tsv(params_tsv)

# Read sample sheet
samplesheet <- read_csv("samplesheets/samplesheet.csv")

# Get samplesheet column names
samplesheet_cols <- list(
  sample = "Sample name",
  rds = "File path",
  min_ncount = "Min. UMI count",
  max_ncount = "Max. UMI count",
  min_nfeature = "Min. gene count",
  max_nfeature = "Max. gene count",
  min_mt_pct = "Min. MT gene proportion (%)",
  max_mt_pct = "Max. MT gene proportion (%)",
  cells_to_remove = "Cells to remove file path",
  res = "Sample clustering resolution",
  multiplet_rate = "Expected multiplet rate"
)

samplesheet_cols_show <- list(
  sample = FALSE,
  rds = FALSE,
  min_ncount = FALSE,
  max_ncount = FALSE,
  min_nfeature = FALSE,
  max_nfeature = FALSE,
  min_mt_pct = FALSE,
  max_mt_pct = FALSE,
  cells_to_remove = FALSE,
  res = FALSE,
  multiplet_rate = FALSE
)

for (col in colnames(samplesheet)) {
  samplesheet_cols_show[[col]] <- TRUE

  if (!(col %in% names(samplesheet_cols))) {
    samplesheet_cols[[col]] <- col
  }
}

samplesheet_cols_show$rds <- FALSE
samplesheet_cols_show$cells_to_remove <- FALSE
samplesheet_cols_final <- samplesheet_cols[unlist(samplesheet_cols_show)]

samplesheet_show <- samplesheet %>%
  select(all_of(names(samplesheet_cols_final)))

# Read custom marker genes if present
custom_marker_genes <- NA
custom_marker_genes_file <- "samplesheets/custom_marker_genes.csv"
if (file.exists(custom_marker_genes_file)) {
  custom_marker_genes <- read_csv(custom_marker_genes_file)
}

# Read comparisons if present
comparisons <- NA
comparisons_file <- "samplesheets/comparisons.csv"
if (file.exists(comparisons_file)) {
  comparisons <- read_csv(comparisons_file)
}

# Get list of pipeline sections that were run
modules_run <- list(
  qc_filter = TRUE,
  doublets = dir.exists("qc_doublets"),
  integration = dir.exists("integration_qc"),
  annotation_cc = dir.exists("annotation_cc"),
  annotation_db = dir.exists("annotation_db"),
  annotation_custom = dir.exists("annotation_custom"),
  annotation_cluster = dir.exists("annotation_clusters"),
  pseudobulk = dir.exists("analysis_pseudo"),
  de = dir.exists("analysis_de"),
  gsea = dir.exists("analysis_gsea"),
  ora = dir.exists("analysis_ora")
)

module_names <- list(
  qc_filter = "Quality control and filtering",
  doublets = "Doublet detection and filtering",
  integration = "Dataset integration",
  annotation_cc = "Cell cycle annotation",
  annotation_db = "Automatic cell type annotation",
  annotation_custom = "Custom cell type annotation",
  annotation_cluster = "Cluster annotation",
  pseudobulk = "Pseudobulking",
  de = "Differential expression",
  gsea = "Gene set enrichment analysis",
  ora = "Over-representation analysis"
)

modules_run_names <- unlist(module_names[unlist(modules_run)])
modules_run_names_str <- paste("- ", modules_run_names, collapse = "\n")
modules_run_names_str <- paste0(modules_run_names_str, "\n")

# Get executive summary info
n_samples <- c(metric = "# Samples", value = nrow(samplesheet_show))
n_cells_filtered <- NULL
n_cells_per_sample <- NULL
n_clusters <- NULL
cluster_annotations <- NULL
n_de_comparisons <- NULL
n_de_genes <- NULL

if (file.exists("qc_doublets.qmd")) {
  doublet_summary_files <- dir(
    "qc_doublets",
    pattern = ".*\\.doublet_summary\\.csv$",
    recursive = TRUE,
    full.names = TRUE
  )
  doublet_summaries <- lapply(doublet_summary_files, function(f) {
    read_csv(f)
  })
  doublet_summary <- bind_rows(doublet_summaries)
  n_cells_filtered <- c(metric = "# Cells (filtered)", value = sum(doublet_summary$n_singlets))
} else if (file.exists("qc_filter.qmd")) {
  filter_summary_files <- dir(
    "qc_filter",
    pattern = ".*\\.hard_filter_summary\\.csv$",
    recursive = TRUE,
    full.names = TRUE
  )
  filter_summaries <- lapply(filter_summary_files, function(f) {
    read_csv(f)
  })
  filter_summary <- bind_rows(filter_summaries)
  n_cells_filtered <- c(metric = "# Cells (filtered)", value = sum(filter_summary$n_cells_post_filtered))
}

if (!is.null(n_cells_filtered)) {
  n_cells_per_sample <- c(metric = "Avg. cells per sample", value = round(as.integer(n_cells_filtered[["value"]]) / as.integer(n_samples[["value"]]), digits = 2))
}

if (file.exists("integration_cluster.qmd")) {
  n_clusters_df <- read_csv(paste0("integration_cluster/", cohort_name, ".integrated.clusters.csv"))
  n_clusters <- c(metric = "# Clusters", value = nrow(n_clusters_df))
}

if (file.exists("annotation.qmd")) {
  cluster_annotations <- read_csv(paste0("annotation_clusters/", cohort_name, ".cluster_cell_type_assignments.csv")) %>%
    group_by(cell_type) %>%
    summarise(n_clusters = n())
}

if (file.exists("analysis_de.qmd")) {
  comparisons_csv <- "samplesheets/comparisons.csv"
  comparisons_df <- read_csv(comparisons_csv)
  n_de_comparisons <- c(metric = "# DE comparisons", value = nrow(comparisons_df))

  sig_de_table <- read_csv(paste0("analysis_de/", cohort_name, ".de.full.csv")) %>%
    filter(sig == "sig")

  n_de_genes <- c(metric = "# DE genes", value = length(unique(sig_de_table$Gene.Symbol)))
}

exec_summary <- bind_rows(n_samples, n_cells_filtered, n_cells_per_sample, n_clusters, n_de_comparisons, n_de_genes)

# Get information for methods section
# TODO: Get values from environment
pipeline_version <- scan("version.txt", character())[1]
nextflow_version <- scan("nextflow.txt", character())[1]
nextflow_command <- scan("cmd.sh", character(), sep = "\n")
nextflow_command <- paste(nextflow_command, collapse = "\n")
nextflow_command <- paste("```bash", nextflow_command, "```", sep = "\n")
software_versions <- read_csv("software.csv")
r_version <- paste(version$major, version$minor, sep = ".")
seurat_version <- as.character(packageVersion("Seurat"))
tidyverse_ersion <- as.character(packageVersion("tidyverse"))
clustree_version <- software_versions$version[match("clustree", software_versions$tool)]
doubletfinder_version <- software_versions$version[match("DoubletFinder", software_versions$tool)]
singler_version <- software_versions$version[match("SingleR", software_versions$tool)]
deseq_version <- software_versions$version[match("DESeq2", software_versions$tool)]
webgestaltr_version <- software_versions$version[match("WebGestaltR", software_versions$tool)]
celldex_version <- software_versions$version[match("celldex", software_versions$tool)]
hard_filter_cols <- c("min_ncount", "max_ncount", "min_nfeature", "max_nfeature")
hard_filter_cols <- hard_filter_cols[hard_filter_cols %in% samplesheet]
hard_filtered <- any(!is.na(samplesheet[hard_filter_cols]))
mt_qc <- !as.logical(params_df$value[match("no_mt", params_df$parameter)])
mt_filter_cols <- c("min_mt_pct", "max_mt_pct")
mt_filter_cols <- mt_filter_cols[mt_filter_cols %in% samplesheet]
mt_filtered <- any(!is.na(samplesheet[mt_filter_cols]))
cluster_algorithm <- str_to_sentence(params_df$value[match("cluster_method", params_df$parameter)])
integration_resolution <- params_df$value[match("integrated_resolution", params_df$parameter)]
annotation_species <- params_df$value[match("species", params_df$parameter)]
annotation_db_param <- params_df$value[match("annotation_db", params_df$parameter)]
if (is.null(annotation_db_param) || is.na(annotation_db_param) || annotation_db_param == "null" || annotation_db_param == "") {
  annotation_db_source <- paste0("the Human Primary Cell Atlas (HPCA) via the R package `celldex` (version ", celldex_version, ")")
} else {
  annotation_db_source <- "a pre-built annotation database"
  annotation_db_source <- paste0(annotation_db_param)
}
annotation_custom_mad_threshold <- params_df$value[match("custom_annotation_mad_threshold", params_df$parameter)]
available_annotation_files <- dir(
  "available_annotations",
  pattern = "\\.txt$",
  recursive = TRUE,
  full.names = TRUE
)
available_annotations <- unlist(sapply(available_annotation_files, scan, character()))
annotation_priority <- c(
  "custom_cell_type",
  "custom_cell_type.max_score",
  "SingleR.annotation",
  "SingleR.hpca_main",
  "SingleR.mouse_main",
  "SingleR.hpca_fine",
  "SingleR.mouse_fine",
  "Phase"
)
annotation_priority <- annotation_priority[annotation_priority %in% available_annotations]
cluster_annotation_param <- params_df$value[match("cluster_annotation", params_df$parameter)]
if (is.null(cluster_annotation_param) || is.na(cluster_annotation_param) || cluster_annotation_param == "null" || cluster_annotation_param == "") {
  cluster_annotation_param <- annotation_priority[1]
}
cluster_annotation_descriptions <- list(
  custom_cell_type = "custom marker gene cell type",
  custom_cell_type.max_score = "custom marker gene cell type",
  SingleR.annotation = "cell type from the SingleR annotation",
  SingleR.hpca_main = "cell type from the SingleR annotation",
  SingleR.mouse_main = "cell type from the SingleR annotation",
  SingleR.hpca_fine = "cell type from the SingleR annotation",
  SingleR.mouse_fine = "cell type from the SingleR annotation",
  Phase = "cell cycle phase"
)
cluster_annotation <- cluster_annotation_descriptions[[cluster_annotation_param]]
cluster_annotation_threshold <- params_df$value[match("cell_type_proportion_threshold", params_df$parameter)]
```

## Analysis summary: `r cohort_name`

The following table summarises the major findings from this analysis:

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

datatable(
  exec_summary,
  colnames = "",
  rownames = FALSE
)
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("annotation.qmd")
#| warning: false
#| error: false
#| message: false
cat("The following annotations were assigned to clusters:\n\n")
```

```{r}
#| echo: false
#| eval: !expr (!is.null(cluster_annotations))
#| warning: false
#| error: false
#| message: false

datatable(
  cluster_annotations,
  colnames = c("Cell type", "# Clusters"),
  rownames = FALSE
)
```

### Sample sheet

The following sample sheet table summarises the metadata and potential filtering thresholds for each sample in the dataset:

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

datatable(
  samplesheet_show,
  colnames = unname(unlist(samplesheet_cols_final))
)
```

The following modules were run for this analysis:

```{r}
#| echo: false
#| results: asis
#| warning: false
#| error: false
#| message: false

cat(modules_run_names_str)
```

### Parameters

The following parameters were set during the pipeline run:

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

datatable(
  params_df,
  colnames = c("Parameter", "Value")
)
```

### Methods

You can use the following text within your methods sections of any manuscripts using the results of this analysis:

::: {.callout-note appearance="minimal"}

#### Bioinformatics and data analysis

Data was processed using scrnavigator-nf `r pipeline_version` (https://github.com/Sydney-Informatics-Hub/scrnavigator-nf), a Nextflow pipeline developed by the Sydney Informatics Hub for analysing single cell RNA sequencing data. The pipeline was executed using version `r nextflow_version` of Nextflow (Di Tommaso, et al., 2017; doi: [10.1038/nbt.3820](http://www.nature.com/nbt/journal/v35/n4/full/nbt.3820.html)) with the following command:

```{r}
#| echo: false
#| results: asis
#| warning: false
#| error: false
#| message: false
cat(nextflow_command)
```

The pipeline is built around the `Seurat` framework (version `r seurat_version`; Satija et al., 2015; doi: [10.1038/nbt.3192](https://doi.org/10.1038/nbt.3192)) and the `R` programming language (version `r r_version`).
```{r}
#| echo: false
#| results: asis
#| warning: false
#| error: false
#| message: false
if (file.exists("qc_filter.qmd") && hard_filtered) {
  cat("Distributions of UMI and gene counts were inspected to select hard filtering thresholds for removal of low-quality cells.\n")
  if (mt_qc && mt_filtered) {
    cat("Mitochondrial gene fractions per cell were also inspected to select hard filtering thresholds for removal of potentailly dying cells.\n")
  }
}
if (file.exists("qc_cluster.qmd")) {
  cat(paste0("Cells were clustered using the ", cluster_algorithm, " algorithm, with clustering resolutions selected using the R package `clustree` (version ", clustree_version, "; Zappia and Oshlack, 2018; doi: [10.1093/gigascience/giy083](https://doi.org/10.1093/gigascience/giy083)).\n"))
}

if (file.exists("qc_doublets.qmd")) {
  cat(paste0("Doublets were detected and removed using the R package `DoubletFinder` (version ", doubletfinder_version, "; McGinnis et al., 2019; doi: [10.1016/j.cels.2019.03.003](https://doi.org/10.1016/j.cels.2019.03.003)).\n"))
}

if (file.exists("integration_qc.qmd")) {
  cat("Samples were merged into a single dataset and integrated to correct for batch effects with the `Seurat` Canonical Correlation Analysis (CCA) algorithm.\n")
}

if (file.exists("integration_cluster.qmd")) {
  cat(paste0("The integrated dataset was re-clustered using the ", cluster_algorithm, " algorithm, with a clustering resolution of ", integration_resolution, " selected using Clustree.\n"))
}

if (file.exists("annotation.qmd")) {
  cat("\nCells were annotated as follows.\n")
  if (dir.exists("annotation_cc")) {
    cat("Cell cycle states were annotated using the `Seurat` built-in function `CellCycleScoring`.\n")
  }

  if (dir.exists("annotation_db")) {
    cat(paste0("Cell types were annotated using the `SingleR` R package (version ", singler_version, "; Aran et al., 2019; doi: [10.1038/s41590-018-0276-y](https://doi.org/10.1038/s41590-018-0276-y)) and ", annotation_db_source, ".\n"))
  }

  if (dir.exists("annotation_custom")) {
    cat("Cells types were annotated using custom marker gene sets and the `Seurat` built-in function `AddModuleScore`.\n")
    if (cluster_annotation_param == "custom_cell_type.max_score") {
      cat(paste0("The cell type with the highest ranking module score was used.\n"))
    } else {
      cat(paste0("The cell type with the highest ranking module score was used if the difference between the top two module scores was greater than ", annotation_custom_mad_threshold, " median absolute deviation (MAD) of all scores.\n"))
      cat("Cells that did not meet this criteria were labelled 'Ambiguous'.\n")
    }
  }

  if (dir.exists("annotation_clusters")) {
    cat(paste0("Clusters were annotated with the most prevalent ", cluster_annotation, " present within them, if that prevalence exceeded ", cluster_annotation_threshold, ".\n"))
    cat("Clusters not meeting this criteria were labelled 'Ambiguous'.\n")
  }
}

if (dir.exists("analysis_pseudo")) {
  cat("\nPseudobulking was performed to aggregate expression values from cells within each sample using `Seurat`'s built-in function `AggregateExpression`.\n")
}

if (dir.exists("analysis_de")) {
  cat(paste0("Differential expression analysis was performed on pseudobulked data using `Seurat`'s built-in function `FindMarkers` and the `DESeq2` R package (version ", deseq_version, "; Love et al., 2014; doi: [10.1186/s13059-014-0550-8](http://dx.doi.org/10.1186/s13059-014-0550-8)).\n"))
}

if (dir.exists("analysis_gsea") || dir.exists("analysis_ora")) {
  cat(paste0("Functional enrichment analysis (FEA) was performed with the `WebGestaltR` R package (version ", webgestaltr_version, "; Elizarraras et al., 2024; doi: [10.1093/nar/gkae456](https://doi.org/10.1093/nar/gkae456)). The following FEA methods were used:\n"))
  if (dir.exists("analysis_gsea")) {
    cat("Gene Set Enrichment Analysis (GSEA).\n")
  }

  if (dir.exists("analysis_ora")) {
    cat("Over-Representation Analysis (ORA).\n")
  }

}
```

:::

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("qc_filter.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Quality control\n\n")
cat("### Initial filtering\n\n")
```

```{r}
#| child: "qc_filter.qmd"
#| eval: !expr file.exists("qc_filter.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("qc_cluster.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Sample clustering\n\n")
```

```{r}
#| child: "qc_cluster.qmd"
#| eval: !expr file.exists("qc_cluster.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("qc_doublets.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Doublet detection and filtering\n\n")
```

```{r}
#| child: "qc_doublets.qmd"
#| eval: !expr file.exists("qc_doublets.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("integration_qc.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Integration\n\n")
cat("### Initial quality control\n\n")
```

```{r}
#| child: "integration_qc.qmd"
#| eval: !expr file.exists("integration_qc.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("integration_cluster.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Clustering\n\n")
```

```{r}
#| child: "integration_cluster.qmd"
#| eval: !expr file.exists("integration_cluster.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("annotation.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Annotation\n\n")
```

```{r}
#| child: "annotation.qmd"
#| eval: !expr file.exists("annotation.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_pseudo.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Analysis\n\n")
cat("### Pseudobulking\n\n")
```

```{r}
#| child: "analysis_pseudo.qmd"
#| eval: !expr file.exists("analysis_pseudo.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_de.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Differential expression analysis\n\n")
```

```{r}
#| child: "analysis_de.qmd"
#| eval: !expr file.exists("analysis_de.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr (file.exists("analysis_gsea.qmd") || file.exists("analysis_ora.qmd"))
#| warning: false
#| error: false
#| message: false

cat("### Functional enrichment analysis (FEA)\n\n")

cat("Differential expression analyses can generate lots of results that can be quite difficult to interpret by themselves.\n")
cat("Functional enrichment analyses are a set of methods designed to take the raw differential expression results and identify patterns of expression that point to altered **biological function**.\n")
cat("There are several such approaches, but the two we apply in this pipeline are **Gene Set Enrichment Analysis** (GSEA) and **Over-Representation Analysis** (ORA).\n\n")
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_gsea.qmd")
#| warning: false
#| error: false
#| message: false
cat("#### Gene Set Enrichment Analysis (GSEA)\n\n")
```

```{r}
#| child: "analysis_gsea.qmd"
#| eval: !expr file.exists("analysis_gsea.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_ora.qmd")
#| warning: false
#| error: false
#| message: false
cat("#### Over-Representation Analysis (ORA)\n\n")
```

```{r}
#| child: "analysis_ora.qmd"
#| eval: !expr file.exists("analysis_ora.qmd")
#| warning: false
#| error: false
#| message: false
```

## Software versions

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

tibble(
  tool = c("R", "Seurat"),
  version = c(r_version, seurat_version)
) %>%
  bind_rows(software_versions) %>%
  datatable(
    colnames = c("Tool", "Version"),
    rownames = FALSE
  )
```