---
title: "scRNAvigator Report"
format:
  html:
    embed-resources: true
---

Thank you for using scRNAvigator!

This report contains a summary of your analysis. Depending on what sections of the pipeline were ran, it will include one or more of the following:

- Quality control metrics and plots
- Doublet detection and filtering results
- Dataset integration results
- Cell annotation results
- Pseudobulking and differential expression results
- Functional enrichment analysis results

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)

# Get cohort name
cohort_name <- scan("cohort_name.txt", character())[1]

# Get parameters TSV file
params_tsv <- "parameters.tsv"
params_df <- read_tsv(params_tsv)

# Read sample sheet
samplesheet <- read_csv("samplesheets/samplesheet.csv")

# Get samplesheet column names
samplesheet_cols <- list(
  sample = "Sample name",
  rds = "File path",
  min_ncount = "Min. UMI count",
  max_ncount = "Max. UMI count",
  min_nfeature = "Min. gene count",
  max_nfeature = "Max. gene count",
  min_mt_pct = "Min. MT gene proportion (%)",
  max_mt_pct = "Max. MT gene proportion (%)",
  cells_to_remove = "Cells to remove file path",
  res = "Sample clustering resolution",
  multiplet_rate = "Expected multiplet rate"
)

samplesheet_cols_show <- list(
  sample = FALSE,
  rds = FALSE,
  min_ncount = FALSE,
  max_ncount = FALSE,
  min_nfeature = FALSE,
  max_nfeature = FALSE,
  min_mt_pct = FALSE,
  max_mt_pct = FALSE,
  cells_to_remove = FALSE,
  res = FALSE,
  multiplet_rate = FALSE
)

for (col in colnames(samplesheet)) {
  samplesheet_cols_show[[col]] <- TRUE

  if (!(col %in% names(samplesheet_cols))) {
    samplesheet_cols[[col]] <- col
  }
}

samplesheet_cols_show$rds <- FALSE
samplesheet_cols_show$cells_to_remove <- FALSE
samplesheet_cols_final <- samplesheet_cols[unlist(samplesheet_cols_show)]

samplesheet_show <- samplesheet %>%
  select(all_of(names(samplesheet_cols_final)))

# Read custom marker genes if present
custom_marker_genes <- NA
custom_marker_genes_file <- "samplesheets/custom_marker_genes.csv"
if (file.exists(custom_marker_genes_file)) {
  custom_marker_genes <- read_csv(custom_marker_genes_file)
}

# Read comparisons if present
comparisons <- NA
comparisons_file <- "samplesheets/comparisons.csv"
if (file.exists(comparisons_file)) {
  comparisons <- read_csv(comparisons_file)
}

# Get list of pipeline sections that were run
modules_run <- list(
  qc_filter = TRUE,
  doublets = dir.exists("qc_doublets"),
  integration = dir.exists("integration_qc"),
  annotation_cc = dir.exists("annotation_cc"),
  annotation_db = dir.exists("annotation_db"),
  annotation_custom = dir.exists("annotation_custom"),
  annotation_cluster = dir.exists("annotation_clusters"),
  pseudobulk = dir.exists("analysis_pseudo"),
  de = dir.exists("analysis_de"),
  gsea = dir.exists("analysis_gsea"),
  ora = dir.exists("analysis_ora")
)

module_names <- list(
  qc_filter = "Quality control and filtering",
  doublets = "Doublet detection and filtering",
  integration = "Dataset integration",
  annotation_cc = "Cell cycle annotation",
  annotation_db = "Automatic cell type annotation",
  annotation_custom = "Custom cell type annotation",
  annotation_cluster = "Cluster annotation",
  pseudobulk = "Pseudobulking",
  de = "Differential expression",
  gsea = "Gene set enrichment analysis",
  ora = "Over-representation analysis"
)

modules_run_names <- unlist(module_names[unlist(modules_run)])
modules_run_names_str <- paste("- ", modules_run_names, collapse = "\n")
modules_run_names_str <- paste0(modules_run_names_str, "\n")

# Get executive summary info
n_samples <- c(metric = "# Samples", value = nrow(samplesheet_show))
n_cells_filtered <- NULL
n_cells_per_sample <- NULL
n_clusters <- NULL
cluster_annotations <- NULL
n_de_comparisons <- NULL
n_de_genes <- NULL

if (file.exists("qc_doublets.qmd")) {
  doublet_summary_files <- dir(
    "qc_doublets",
    pattern = ".*\\.doublet_summary\\.csv$",
    recursive = TRUE,
    full.names = TRUE
  )
  doublet_summaries <- lapply(doublet_summary_files, function(f) {
    read_csv(f)
  })
  doublet_summary <- bind_rows(doublet_summaries)
  n_cells_filtered <- c(metric = "# Cells (filtered)", value = sum(doublet_summary$n_singlets))
} else if (file.exists("qc_filter.qmd")) {
  filter_summary_files <- dir(
    "qc_filter",
    pattern = ".*\\.hard_filter_summary\\.csv$",
    recursive = TRUE,
    full.names = TRUE
  )
  filter_summaries <- lapply(filter_summary_files, function(f) {
    read_csv(f)
  })
  filter_summary <- bind_rows(filter_summaries)
  n_cells_filtered <- c(metric = "# Cells (filtered)", value = sum(filter_summary$n_cells_post_filtered))
}

if (!is.null(n_cells_filtered)) {
  n_cells_per_sample <- c(metric = "Avg. cells per sample", value = round(as.integer(n_cells_filtered[["value"]]) / as.integer(n_samples[["value"]]), digits = 2))
}

if (file.exists("integration_cluster.qmd")) {
  n_clusters_df <- read_csv(paste0("integration_cluster/", cohort_name, ".integrated.clusters.csv"))
  n_clusters <- c(metric = "# Clusters", value = nrow(n_clusters_df))
}

if (file.exists("annotation.qmd")) {
  cluster_annotations <- read_csv(paste0("annotation_clusters/", cohort_name, ".cluster_cell_type_assignments.csv")) %>%
    group_by(cell_type) %>%
    summarise(n_clusters = n())
}

if (file.exists("analysis_de.qmd")) {
  comparisons_csv <- "samplesheets/comparisons.csv"
  comparisons_df <- read_csv(comparisons_csv)
  n_de_comparisons <- c(metric = "# DE comparisons", value = nrow(comparisons_df))

  sig_de_table <- read_csv(paste0("analysis_de/", cohort_name, ".de.full.csv")) %>%
    filter(sig == "sig")

  n_de_genes <- c(metric = "# DE genes", value = length(unique(sig_de_table$Gene.Symbol)))
}

exec_summary <- bind_rows(n_samples, n_cells_filtered, n_cells_per_sample, n_clusters, n_de_comparisons, n_de_genes)
```

## Analysis summary: `r cohort_name`

The following table summarises the major findings from this analysis:

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

datatable(
  exec_summary,
  colnames = "",
  rownames = FALSE
)
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("annotation.qmd")
#| warning: false
#| error: false
#| message: false
cat("The following annotations were assigned to clusters:\n\n")
```

```{r}
#| echo: false
#| eval: !expr (!is.null(cluster_annotations))
#| warning: false
#| error: false
#| message: false

datatable(
  cluster_annotations,
  colnames = c("Cell type", "# Clusters"),
  rownames = FALSE
)
```

### Sample sheet

The following sample sheet table summarises the metadata and potential filtering thresholds for each sample in the dataset:

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

datatable(
  samplesheet_show,
  colnames = unname(unlist(samplesheet_cols_final))
)
```

The following modules were run for this analysis:

```{r}
#| echo: false
#| results: asis
#| warning: false
#| error: false
#| message: false

cat(modules_run_names_str)
```

### Parameters

The following parameters were set during the pipeline run:

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false

datatable(
  params_df,
  colnames = c("Parameter", "Value")
)
```

### Methods

TODO

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("qc_filter.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Quality control\n\n")
cat("### Initial filtering\n\n")
```

```{r}
#| child: "qc_filter.qmd"
#| eval: !expr file.exists("qc_filter.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("qc_cluster.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Clustering\n\n")
```

```{r}
#| child: "qc_cluster.qmd"
#| eval: !expr file.exists("qc_cluster.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("qc_doublets.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Doublet detection and filtering\n\n")
```

```{r}
#| child: "qc_doublets.qmd"
#| eval: !expr file.exists("qc_doublets.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("integration_qc.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Integration\n\n")
cat("### Initial quality control\n\n")
```

```{r}
#| child: "integration_qc.qmd"
#| eval: !expr file.exists("integration_qc.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("integration_cluster.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Clustering\n\n")
```

```{r}
#| child: "integration_cluster.qmd"
#| eval: !expr file.exists("integration_cluster.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("annotation.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Annotation\n\n")
```

```{r}
#| child: "annotation.qmd"
#| eval: !expr file.exists("annotation.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_pseudo.qmd")
#| warning: false
#| error: false
#| message: false
cat("## Analysis\n\n")
cat("### Pseudobulking\n\n")
```

```{r}
#| child: "analysis_pseudo.qmd"
#| eval: !expr file.exists("analysis_pseudo.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_de.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Differential expression analysis\n\n")
```

```{r}
#| child: "analysis_de.qmd"
#| eval: !expr file.exists("analysis_de.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_gsea.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Functional enrichment analysis: GSEA\n\n")
```

```{r}
#| child: "analysis_gsea.qmd"
#| eval: !expr file.exists("analysis_gsea.qmd")
#| warning: false
#| error: false
#| message: false
```

```{r}
#| echo: false
#| results: asis
#| eval: !expr file.exists("analysis_ora.qmd")
#| warning: false
#| error: false
#| message: false
cat("### Functional enrichment analysis: ORA\n\n")
```

```{r}
#| child: "analysis_ora.qmd"
#| eval: !expr file.exists("analysis_ora.qmd")
#| warning: false
#| error: false
#| message: false
```
