---
title: "Integration: Clustering"
format: html
---

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)
```

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
# Integration cluster
# Get UMAP
umap_file <- dir(
  "integration_cluster",
  pattern = "^.*\\.umap\\.integrated\\.clusters\\.png$",
  recursive = TRUE,
  full.names = TRUE
)[1]

# Get cohort ID
cohort_id <- gsub("^integration_cluster\\/(.*)\\.umap\\.integrated\\.clusters\\.png$", "\\1", umap_file)

# Get feature-count plot
fc_plot_file <- paste0("integration_cluster/", cohort_id, ".feature_count_plot.integrated.cluster.png")

# Get clustree plot
clustree_file <- paste0("integration_cluster/", cohort_id, ".clustree.png")

# Get integration resolution
integration_res_file <- "integration_resolution.txt"
integration_res <- as.numeric(scan(integration_res_file, character())[1])
```

This section shows the clustering results for your dataset post-integration. As before with the [per-sample clustering analysis](#sample-clustering), the dataset was clustered at several resolutions (outlined below). A cluster tree plot is shown below to aid you in selecting the best resolution for your dataset. Recall that we want to identify a resolution where the clusters remain relatively stable from one resolution to the next, i.e. most clusters have a single, opaque arrow incoming rather than multiple faint incoming arrows. By default, the pipeline will choose a value of `1`, **but often this will need to be altered to suit the dataset**. For your dataset, a value of **`r integration_res`** was selected. The UMAP and count vs. feature plots below will display the clustering results for your chosen resolution.

#### Integrated clusters UMAP

These UMAP plots show your integrated dataset, colour-coded by either the sample of origin ("orig.ident") or the cluster each cell was assigned to in the chosen integration clustering resolution ("SCT_snn_res.*").

```{r}
#| results: asis
#| echo: false
cat(paste0('<img src="', umap_file, '" alt="integrated clusters umap: ', cohort_id, '" width="100%">\n\n'))
```

#### Feature-count plot of integrated dataset

This plot shows the count vs. feature plots for each cluster at your chosen clustering resolution.

```{r}
#| results: asis
#| echo: false
cat(paste0('<img src="', fc_plot_file, '" alt="feature-count plot of integrated dataset: ', cohort_id, '" width="100%">\n\n'))
```

#### Clustree plot for integrated dataset

This cluster tree shows how cells move between clusters from one resolution to the next. Stable resolutions are where most clusters have a single, dark incoming arrow, meaning that few cells moved between clusters when moving from the previous resolution.

```{r}
#| results: asis
#| echo: false
cat(paste0('<img src="', clustree_file, '" alt="clustree plot for integrated dataset: ', cohort_id, '" width="100%">\n\n'))
```
