---
title: "Quality Control: Doublet Detection"
format: html
---

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)
```

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
# QC doublets
# Get doublet summary files
doublet_summary_files <- dir(
  "qc_doublets",
  pattern = ".*\\.doublet_summary\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)
doublet_summaries <- lapply(doublet_summary_files, function(f) {
  read_csv(f)
})
doublet_summary <- bind_rows(doublet_summaries) %>%
  mutate(prop_doublets = (n_doublets / (n_singlets + n_doublets)) * 100) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

# Get sample names and map them to directory names
samples <- sapply(doublet_summary_files, function(s) {
  gsub("^qc_doublets\\/.*\\/(.*)\\.doublet_summary\\.csv$", "\\1", s)
})
names(samples) <- sapply(names(samples), function(s) {
  gsub("^qc_doublets\\/(.*)\\/.*$", "\\1", s)
})
sample_map <- setNames(names(samples), samples)

# Get doublet per cluster plots
doublet_per_cluster_plots <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  dir(
    paste("qc_doublets", i, sep = "/"),
    pattern = paste(s, "doublets_per_cluster", ".*", "png$", sep = "\\."),
    full.names = TRUE
  )
})
# Get primary resolution per sample
primary_cluster_res <- sapply(doublet_per_cluster_plots,function(f) {
  as.numeric(gsub("^.*\\.doublets_per_cluster\\.(.*)\\.png$", "\\1", f))
})

# Get doublet proportion per cluster plots
doublet_prop_per_cluster_plots <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  res <- primary_cluster_res[[s]]
  paste("qc_doublets", i, paste(s, "doublet_proportions_per_cluster", res, "png", sep = "."), sep = "/")
})

# Get doublet UMAPs
doublet_umap_plots <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  res <- primary_cluster_res[[s]]
  paste("qc_doublets", i, paste(s, "doublet_umap", res, "png", sep = "."), sep = "/")
})

# Get clustree plot
clustree_plot_files <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  paste("qc_doublets", i, paste(s, "clustree", "png", sep = "."), sep = "/")
})

# Get feature-count plots for each clustering resolution
feature_count_plot_files <- lapply(sample_map, function(i) {
  s <- samples[[i]]
  dir(
    paste("qc_doublets", i, sep = "/"),
    pattern = paste(s, "feature_count_plot", "cluster", ".*","png", sep = "\\."),
    full.names = TRUE
  )
})

# Get cluster resolutions for each sample (should be the same)
cluster_res <- lapply(feature_count_plot_files, function(f) {
  as.numeric(gsub("^.*\\.feature_count_plot\\.cluster\\.(.*)\\.png$", "\\1", f))
})
for (n in names(feature_count_plot_files)) {
  names(feature_count_plot_files[[n]]) <- cluster_res[[n]]
}
```

Doublets are cases of two cells being captured together during the sample preparation phase, causing both cells to be barcoded with the same DNA barcode. This results in the individual cells' expression profiles being merged together. These doublets (or multiplets in the case of three or more cells being captured together) don't represent real cells and will confound donwstream analyses. As such, they should be removed. We have used the R package `DoubletFinder` to identify barcodes that are likely doublets and removed them from the dataset.

Below you will find several plots for each sample illustrating the doublets identified. At the top are two bar graphs showing the number and proportion of doublets identified per cluster, respectively. You will also find a UMAP plot of your single cell data with doublets highlighted in red to show which clusters were affected. Below these plots you will find clustering plots similar to [those above](#sample-clustering), including a cluster tree plot and multiple count vs. feature plots for each cluster of cells at each clustering resolution. This clustering analysis has been performed after doublet removal and demonstrate the effect of doublet removal on the clustering results.

Finally, below the plots you will find a summary table with the number and proportion of doublets identified per sample.

:::panel-tabset

```{r}
#| results: asis
# Display clustree plots and per-resolution feature-count plots
purrr::walk(
  sort(samples),
  function(n) {
    prim_res <- primary_cluster_res[[n]]
    prim_res_str <- paste0("(resolution: ", prim_res, ")")
    p_doublets <- doublet_per_cluster_plots[[n]]
    p_doublet_props <- doublet_prop_per_cluster_plots[[n]]
    p_umap <- doublet_umap_plots[[n]]
    p_clustree <- clustree_plot_files[[n]]
    p_fc_plot_list <- feature_count_plot_files[[n]]
    cat(paste("####", n, "\n\n"))
    cat(paste("#####", "Doublets per cluster", prim_res_str, "\n\n"))
    cat(paste0('<img src="', p_doublets, '" alt="doublet per cluster plot: ', n, '; res: ', prim_res, '" width="100%">\n\n'))
    cat(paste("#####", "Doublet proportions per cluster", prim_res_str, "\n\n"))
    cat(paste0('<img src="', p_doublet_props, '" alt="doublet proportions per cluster plot: ', n, '; res: ', prim_res, '" width="100%">\n\n'))
    cat(paste("#####", "Doublet UMAP", prim_res_str, "\n\n"))
    cat(paste0('<img src="', p_umap, '" alt="doublet UMAP: ', n, '; res: ', prim_res, '" width="100%">\n\n'))
    cat(paste("#####", "Clustree plot", "\n\n"))
    cat(paste0('<img src="', p_clustree, '" alt="clustree plot: ', n, '" width="100%">\n\n'))
    cat(paste("#####", "Per-cluster feature-count plots", "\n\n"))
    cat(":::panel-tabset\n\n")
    for (res in sort(names(p_fc_plot_list))) {
      p_fc_plot <- p_fc_plot_list[[res]]
      cat(paste("######", res, "\n\n"))
      cat(paste("Cluster resolution:", res, "\n\n"))
      cat(paste0('<img src="', p_fc_plot, '" alt="Per-cluster feature-count plot: ', n, '; res: ', res, '" width="100%">\n\n'))
    }
    cat(":::\n\n")
  }
)
```

:::

#### Doublet filtering summary

```{r}
#| echo: false
datatable(
  doublet_summary,
  colnames = c("Sample", "# Singlets", "# Doublets", "Doublet rate (%)")
)
```
