---
title: "Quality Control: Initial Filtering"
format: html
---

```{r}
#| echo: false
#| results: hide
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(DT)
```

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
# QC filter
# Get filter summaries
filter_summary_files <- dir(
  "qc_filter",
  pattern = ".*\\.hard_filter_summary\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)
filter_summaries <- lapply(filter_summary_files, function(f) {
  read_csv(f)
})
filter_summary <- bind_rows(filter_summaries) %>%
  mutate(prop_filtered = prop_filtered * 100) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

# Get sample names and map them to directory names
samples <- sapply(filter_summary_files, function(s) {
  gsub("^qc_filter\\/.*\\/(.*)\\.hard_filter_summary\\.csv$", "\\1", s)
})
names(samples) <- sapply(names(samples), function(s) {
  gsub("^qc_filter\\/(.*)\\/.*$", "\\1", s)
})
sample_map <- setNames(names(samples), samples)

# Get count distribution graphs
cdist_log_graph_files <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  paste("qc_filter", i, paste(s, "count_distributions.log.png", sep = "."), sep = "/")
})
cdist_graph_files <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  paste("qc_filter", i, paste(s, "count_distributions.png", sep = "."), sep = "/")
})

# Get pre- and post-filtering plots
pre_filter_plots <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  paste("qc_filter", i, paste(s, "feature_count_plot.pre_filtered.png", sep = "."), sep = "/")
})
post_filter_plots <- sapply(sample_map, function(i) {
  s <- samples[[i]]
  paste("qc_filter", i, paste(s, "feature_count_plot.post_filtered.png", sep = "."), sep = "/")
})

# Read sample sheet
samplesheet <- read_csv("samplesheets/samplesheet.csv")

filtering_params <- samplesheet %>%
  select(any_of(c(
    "sample",
    "min_ncount",
    "max_ncount",
    "min_nfeature",
    "max_nfeature",
    "min_mt_pct",
    "max_mt_pct",
    "cells_to_remove"
  )))

is_filtered <- !all(is.na(filtering_params) | filtering_params == "")
```

The plots on this page show the distributions of UMI counts (called `nCount_RNA`) and unique gene counts (called `nFeature_RNA`) per cell, as well as the mitochondiral gene percentages (`percent.mt`) per cell. These three values are plotted in the "Features vs counts" scatter plots towards the bottom of the page. Ideally, you want to see a roughly linear relationship between these values, up to a plateau point where greater sequencing depth no longer captures additional genes. Poor quality or dying cells may appear as additional clusters of cells towards the bottom-right corners of the plots. Additionally, high mitochondrial gene percentages may be indicative of dying cells, although some cell types may show high mitochondrial gene percentages even when healthy.

The table at the bottom of the page summarises the number of cells in each sample. If per-sample filtering was performed, the pre- and post-filter cell counts, as well as the percentage of cells filtered per sample, are shown in this table as well.

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
#| results: asis
#| eval: !expr is_filtered

markdown_text <- "#### Filtering thresholds

The samples in the dataset were filtered according to the following thresholds:

"

cat(markdown_text)
```

```{r}
#| echo: false
#| warning: false
#| error: false
#| message: false
#| eval: !expr is_filtered

filtering_params_cols <- c(
  sample = "Sample name",
  min_ncount = "Min. UMI count",
  max_ncount = "Max. UMI count",
  min_nfeature = "Min. gene count",
  max_nfeature = "Max. gene count",
  min_mt_pct = "Min. MT gene proportion (%)",
  max_mt_pct = "Max. MT gene proportion (%)",
  cells_to_remove = "Cells to remove file path"
)

filtering_params_non_null <- filtering_params %>%
  select(where(~ !all(is.na(.x) | .x == ""))) %>%
  filter(!if_all(everything(), ~ is.na(.x) | .x == ""))

filtering_params_non_null_cols <- filtering_params_cols[colnames(filtering_params_non_null)]

datatable(filtering_params_non_null, colnames = unname(filtering_params_non_null_cols))
```

:::panel-tabset

```{r}
#| results: asis
# Display count distribution plots
purrr::walk(
  sort(samples),
  function(n) {
    p <- cdist_graph_files[[n]]
    p_log <- cdist_log_graph_files[[n]]
    f_pre <- pre_filter_plots[[n]]
    f_post <- post_filter_plots[[n]]
    cat(paste("####", n, "\n\n"))
    cat(paste("#####", "Count distributions", "\n\n"))
    cat(paste0('<img src="', p, '" alt="count distribution plot: ', n, '" width="100%">\n\n'))
    cat(paste0('<img src="', p_log, '" alt="log-count distribution plot: ', n, '" width="100%">\n\n'))
    cat(paste("#####", "Features vs counts - pre-filter", "\n\n"))
    cat(paste0('<img src="', f_pre, '" alt="pre-filter plot: ', n, '" width="100%">\n\n'))
    cat(paste("#####", "Features vs counts - post-filter", "\n\n"))
    cat(paste0('<img src="', f_post, '" alt="post-filter plot: ', n, '" width="100%">\n\n'))
  }
)
```

:::

#### Filtering summary

```{r}
#| echo: false
datatable(
  filter_summary,
  colnames = c("Sample", "# Cells Pre-Filtering", "# Cells Post-Filtering", "% Filtered")
)
```
